---
title: "Exporing the noncancer cohort results"
author: "Elsie Horne"
date: "17/02/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(glue)
cohort <- "mrna"
source(here("analysis", "design.R"))
```

# Background

In our investigation of incremental vaccine effectiveness for three vs two doses, estimated hazard ratios (HRs) against non-COVID-19 death were almost as small as those against COVID-19 death (Figure xx).
We hypothesised that a reason for this was that many of the non-COVID-19 deaths were either COVID-19 deaths that had not been coded as such, or deaths that were indirectly linked to COVID due to health complications following COVID-19.
To investigate this further, we estimated HRs for CVD-related non-COVID deaths and cancer-related non-COVID-19 deaths, on the basis that cancer-related non-COVID-19 deaths were less likely to be linked to complications of COVID-19 compared to CVD-related non-COVID-19 deaths.
However, estimated HRs for cancer-related non-COVID-19 death were as small, if not smaller than those for CVD-related non-COVID-19 death (Figure xx).
In particular, the estimated HRs were much smaller for cancer-related non-COVID-19 death in the period 0-14 days after third dose (xxx and xxx for cancer-related and CVD-related deaths respectively).
We hypothesised that a reason for this could be that people who are dying of cancer do not receive a third dose, so are underrepresented in the three dose group and overrepresented in the two dose group.
To investigate this we repeated the analysis in a "non-cancer" cohort, obtained by excluding people with either of the following recorded in the 5 years prior to the trial start date: 
- a hospital admission with a cancer-related ICD10 code in any field
- a cancer-related SNOMED code in their primary care record 

This report explores the results of the analysis in the non-cancer cohort.

## Study population
```{r }
# read data_matched
data_matched <- read_rds(here("output", cohort, "match", "data_matched.rds")) 

data <- data_matched %>%
    left_join(
      # data for cancer flag
      arrow::read_feather(here("output", cohort, "cancer", "extract", "input_cancer.feather")) %>%
        mutate(across(ends_with("_date"), as.Date)), 
      by = "patient_id"
    ) %>%
    mutate(
      # flag for cancer code in previous 5 years
      cancer_hospitalisation_5yrs = case_when(
        !is.na(cancer_hospitalisation_before_date) & 
          trial_date - lubridate::years(5) <= cancer_hospitalisation_before_date
        ~ TRUE,
        !is.na(cancer_hospitalisation_after_date) &
          cancer_hospitalisation_after_date < trial_date
        ~ TRUE,
        TRUE ~ FALSE
      ),
      cancer_primarycare_5yrs = case_when(
        !is.na(cancer_primarycare_before_date) & 
          trial_date - lubridate::years(5) <= cancer_primarycare_before_date
        ~ TRUE,
        !is.na(cancer_primarycare_after_date) &
          cancer_primarycare_after_date < trial_date
        ~ TRUE,
        TRUE ~ FALSE
      ),
      noncancer = !(cancer_hospitalisation_5yrs | cancer_primarycare_5yrs)
    ) %>%
  left_join(
      # non-covid death cause data
      arrow::read_feather(here("output", cohort, "noncoviddeathcause", "extract", "input_noncoviddeathcause.feather")) %>%
        mutate(across(ends_with("_date"), as.Date)), 
      by = "patient_id"
    ) %>%
    # cvd or cancer deaths must be non-covid
    mutate(across(
      c(cvddeath_date, cancerdeath_date),
      ~if_else(
        !is.na(.x) & is.na(coviddeath_date), 
        # use noncoviddeath_date in there is any mismatch due to cvddeath_date and 
        # cancerdeath_date being extracted on a different date
        noncoviddeath_date,
        as.Date(NA_character_)
      )
    )) %>%
  select(patient_id, trial_date, dereg_date, treated, noncancer, starts_with("cancer")) 

n_noncancer <- data %>% filter(noncancer) %>% nrow()
n_main <- data %>% nrow()
pct_noncancer <- round(100*n_noncancer/n_main,2)
```

The non-cancer cohort comprised `r scales::comma(n_noncancer, accuracy=1)` individuals, (`r pct_noncancer`% of the `r scales::comma(n_main, accuracy=1)` individuals in the main cohort).

```{r}
km_contrasts_rounded <- read_csv(
  here("output", cohort, "models", "combined", "km_contrasts_rounded.csv")
  )

km_data <- km_contrasts_rounded %>%
  filter(
    subgroup %in% c("all", "noncancer"),
    filename == "overall",
    variant_option == "ignore",
    outcome == "cancerdeath"
    ) %>%
  select(subgroup, ends_with(c("_0", "_1"))) %>%
  pivot_longer(
    cols = ends_with(c("_0", "_1")),
    names_pattern = "(.*)_(\\d)",
    names_to = c(".value", "treated")
  ) %>%
  mutate(personyears = persontime/365.25) %>%
  select(-treated) %>%
  group_by(subgroup) %>%
  summarise(across(c(personyears, n.atrisk, n.event), sum)) %>%
  ungroup() %>%
  mutate(inc = round(1000*n.event/personyears, 3)) %>%
  mutate(across(c(personyears, n.event), scales::comma, accuracy=1))

vals_all <- km_data %>% filter(subgroup=="all")
vals_noncancer <- km_data %>% filter(subgroup=="noncancer")

```

There were `r vals_noncancer %>% pull(n.event)` cancer-related deaths in `r  vals_noncancer %>% pull(personyears)` person-years in the non-cancer cohort (incidence rate `r vals_noncancer %>% pull(inc)` per 1,000), compared to `r vals_all %>% pull(n.event)` in `r  vals_all %>% pull(personyears)` in the main cohort (`r vals_all %>% pull(inc)`).

The following chunk calculates the number of cancer-related deaths in the non-cancer cohort, for which the individual had a cancer code recorded after the trial start date:

```{r}
data %>%
  # noncancer cohort with cancer death before dereg or end of follow-up
  filter(
    noncancer,
    !is.na(cancerdeath_date),
    (is.na(dereg_date) | cancerdeath_date < dereg_date),
    cancerdeath_date <= trial_date + lubridate::days(maxfup)
    ) %>%
  mutate(
    cancer_hospitalisation = trial_date <= cancer_hospitalisation_after_date,
    cancer_primarycare = trial_date <= cancer_primarycare_after_date
  ) %>%
  # mutate(across(c(cancer_hospitalisation, cancer_primarycare), replace_na, replace = FALSE)) %>%
  group_by(cancer_hospitalisation, cancer_primarycare) %>%
  count()
```



